---
title: "single cell basic data processing"
author: "Kent Riemondy RBI"
date: "`R Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: false
    theme: cosmo
    highlight: tango
    fig_caption: true
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
```


```{r libs, message=FALSE, warning=FALSE, echo=FALSE}
source("../R/utils.R")
library(future)
plan(strategy = "multicore", workers = 4)

options(future.globals.maxSize = 2 * 1024 ^ 3)

proj_dir <- here()
data_dir <- file.path(proj_dir, "data", "cellranger")
doc_dir <- file.path(proj_dir, "docs")

fig_dir <- "figs"
mkrs_dir <- "markers"

walk(c(fig_dir,mkrs_dir), dir.create, showWarnings = F)
```

# Experiment Summary


```{r get_data}
samples <- c("1_control", "2_mutant")

sample_paths <- file.path(data_dir,
                          samples,
                          "outs", 
                          "filtered_feature_bc_matrix")

names(sample_paths) <- samples

mat <- Read10X(sample_paths)
```


## General QC for library prep  {.tabset}

```{r }

metrics_paths <- file.path(data_dir,
                          samples,
                          "outs", 
                          "metrics_summary.csv")

names(metrics_paths) <- samples

mapping_dat <- map_dfr(metrics_paths, read_csv, .id = "sample")

clean_up_metadata <- function(metrics_summary) {
  metrics_summary <- mutate_all(metrics_summary, str_remove, "%$")
  metrics_summary <- mutate_at(metrics_summary, .vars= 2:ncol(metrics_summary), as.numeric)
  metrics_summary
}

mapping_dat <- clean_up_metadata(mapping_dat)

metrics <- colnames(mapping_dat)[2:ncol(mapping_dat)]

mapping_dat <- mapping_dat %>% 
  gather(metric, value, -sample) %>% 
  mutate(sample = factor(sample, levels = unique(sample)))

p <- map(metrics, 
    function(x) {
    filter(mapping_dat, metric == x) %>% 
          ggplot(aes(sample, value)) +
            geom_point(aes(color = sample)) +
        scale_color_brewer(palette = "Paired") + 
        labs(y = x, 
             x = "") +
        theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
}) 
```

```{r, results ='asis'}

for(i in seq_along(p)){
  cat('\n### ', metrics[i], '\n')
  print(p[i])
  cat('\n')
}
```


## Preprocessing {.tabset}

```{r create_seurat, message = F, results = 'hide', warning = F}
sobj <- CreateSeuratObject(
  mat,
  min.cells = 3,
  min.features = 200,
  names.delim = "_",
  names.field = 2
)

rm(mat)
gc()


sobj <- PercentageFeatureSet(sobj, pattern = "^mt-", col.name = "percent.mt")

sobj$sample <- str_split(sobj$orig.ident, "_", simplify = TRUE) %>% 
  .[, 1]

dir.create("objects", showWarnings = FALSE)

saveRDS(sobj, file.path("objects", "unfiltered.rds"))
```

### Percent Mitochondrial UMIs 

```{r}
plot_violin(sobj@meta.data, 
            "orig.ident",
            "percent.mt",
            .fill = "orig.ident") +
  labs(x = "", y = "Percent UMIs from Mitochondria")
```


### # of genes detected 

```{r}
plot_violin(sobj@meta.data, 
            "orig.ident",
            "nFeature_RNA", .fill = "orig.ident") +
  labs(x = "", y = "# of genes per cell")
```


### # of UMIs detected

```{r}
plot_violin(sobj@meta.data, "orig.ident", "nCount_RNA", .fill = "orig.ident") +
  labs(x = "", y = "# of UMIs") 
```

### Table of mitochondrial proportions per sample

```{r}
sobj@meta.data %>% 
  group_by(orig.ident) %>% 
  summarize(median_percent_mito = median(percent.mt), 
            mean_percent_mito = mean(percent.mt)) %>% 
  arrange(desc(median_percent_mito))

```


## Relationship between UMIs and % mitochondria {.tabset}

### All samples
```{r, fig.width = 12, fig.height = 8}
sample_names <- as.character(unique(sobj@meta.data$orig.ident))
per_sample <- map(sample_names, ~filter(sobj@meta.data, 
                                        orig.ident == .x))
p <- list()
for(i in seq_along(per_sample)){
  .col <- discrete_palette_default[i]
  p[[i]] <- ggplot(per_sample[[i]], aes(nCount_RNA, percent.mt)) +
        geom_point(aes(color = orig.ident)) +
        scale_color_manual(values = .col)
}

plot_grid(plotlist= p, nrow = 1, ncol = 2)
```

```{r, results ='asis'}
# generate tab with individual plot programmatically 
# see https://stackoverflow.com/questions/43752095/programmatically-insert-header-and-plot-in-same-code-chunk-with-r-markdown-using?noredirect=1&lq=1

for(i in seq_along(per_sample)){
  .col <- discrete_palette_default[i]
  cat('\n### ', sample_names[i], '\n')
  p <- ggplot(per_sample[[i]], aes(nCount_RNA, percent.mt)) +
        geom_point(aes(color = orig.ident)) +
        scale_color_manual(values = .col)
  print(p)
  cat('\n')
}
```


## Filter cells and samples.

Suggest removing cells with > 30% mitochondrial UMIs, > 40K UMIs. 
```{r}
sobj <- subset(sobj, subset = percent.mt < 30)
sobj <- subset(sobj, subset = nCount_RNA < 50000)
sobj <- subset(sobj, subset = nCount_RNA > 500)

```



## Normalize and embed into 2D with UMAP

```{r}
sobj <- NormalizeData(sobj, verbose = FALSE)
sobj <- FindVariableFeatures(sobj, 
                             selection.method = "vst", 
                             nfeatures = 3000, 
                             verbose = FALSE)

sobj <- ScaleData(sobj, 
                  verbose = TRUE)

sobj <- RunPCA(sobj, npcs = 40, verbose = FALSE)
sobj <- RunUMAP(sobj, 
                reduction = "pca", 
                dims = 1:40, 
                min.dist = 0.3, 
                n.neighbors = 30L)

plot_feature(sobj, "sample", embedding = "umap", pt_alpha = 0.5)


sobj <- FindNeighbors(sobj, reduction = "pca", dims = 1:40)
sobj <- FindClusters(sobj, resolution = 0.5)

plot_feature(sobj, "seurat_clusters", embedding = "umap", pt_alpha = 0.5)
sobj$clusters <- sobj$seurat_clusters
sobj$seurat_clusters <- NULL
sobj$RNA_snn_res.0.5 <- NULL

mkrs <- FindAllMarkers(sobj, only.pos = TRUE)

#write_tsv(mkrs, file.path(mkrs_dir, "per_cluster_markers.tsv"))

```

Add cell cycle stage

```{r, fig.width = 12}
sobj <- CellCycleScoring(sobj,
                         cc.genes$s.genes,
                         cc.genes$g2m.genes)

cc_scores <- c(
   "Phase",
  "S.Score",
  "G2M.Score"
)

plts <- map(cc_scores, ~plot_feature(sobj, .x, embedding = "umap"))
plt <- plot_grid(plotlist = plts, nrow = 1)
plt

```

Assign cell_types with clustifyr

```{r}
library(clustifyr)
library(ComplexHeatmap)
tm_average <- readRDS(here("data/tabula_muris/TM_avg_expr.rds"))

mdata <- get_metadata(sobj)

res <- clustify(sobj@assays$RNA@data, 
                tm_average, 
                query_genes = sobj@assays$RNA@var.features,
                metadata = mdata, 
                cluster_col = "clusters", 
                compute_method = "spearman")

hmap <- Heatmap(t(res), 
                viridis::viridis(256), 
                "Spearman\ncorrelation",
                row_title = "Cell types from Tabula Muris",
                column_title = "Clusters")

hmap

top_cell_types <- cor_to_call(res)

sobj@meta.data <- sobj@meta.data %>% 
  tibble::rownames_to_column("cell") %>% 
  left_join(top_cell_types, 
            by = c("clusters" = "cluster")) %>% 
  dplyr::rename(tabula_muris_cell_type = type,
                cor_value = r) %>% 
  tibble::column_to_rownames("cell")

plot_umap(sobj, "tabula_muris_cell_type")
```

Write out metadata
```{r} 
mdata_out <- get_metadata(sobj)
dir.create("tables")
outname <- file.path("tables",
                     paste0("metadata_", 
                            format(Sys.Date(), "%Y_%m_%d"),
                            ".tsv.gz"))
#write_tsv(mdata_out, outname)
```

## save merged object
```{r}
#saveRDS(sobj, file.path("objects", "sobj.rds"))
```